---
title: TidyTuesday Week 42
author: Alex
date: '2021-10-12'
slug: ttw42
categories: []
tags: 
  - R
  - ggplot
  - TidyTuesday
---

### TidyTuesday

`TidyTuesday` is a weekly data project put on by the R4DS Learning Community. For more info, see [here](https://github.com/rfordatascience/tidytuesday/blob/master/README.md).

### Setup

```{r Setup}
library(tidyverse)
library(alex) #custom/personal theme package
library(janitor)

fish = read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-12/fish-stocks-within-sustainable-levels.csv') %>% 
  clean_names()
```

### Data Manipulation

I was interested in putting something together similar to a pyramid chart. While shares that add up to 100 aren't really pyramids in shape, bar charts that go off in both directions are interesting ways of viewing data when we only have two categories to follow. While I am by no means an expert in visualization techniques, it seems useful to be able to follow either category (`overexploted` and `sustainable`) without needing to interpret the other from context.

To do so, I'll need to use factors when I plot my data - I've found that `R` struggles when plotting a continuous variable on the y axis, at least in this context. I'll also need to pivot the data into a long form time series from its current wide form, and fill in the blank years with zero, so that I don't miss years in the factors because there is no data for them. I'll filter my data to only use data at a worldwide level, as well.

```{r Manipulation}
fish = fish %>%
  filter(code == "OWID_WRL") %>%
  rename(overexploited = share_of_fish_stocks_that_are_overexploited,
         sustainable = share_of_fish_stocks_within_biologically_sustainable_levels_fao_2020) %>%
  #mutate(sustainable = -1*sustainable) %>%
  select(year, sustainable, overexploited) %>%
  pivot_longer(!year, names_to = "category", values_to = "share")

blank_years = c(2016,2014,2012,2010,2007,2005,2003,2002,2001,1999,1998,1996,1994,1993,1991,1988,1986,
                1984,1982,1980,1976,1975)
blank_years = cbind(blank_years, rep(0,22), rep(0,22))
colnames(blank_years) = c("year", "category", "share")
fish = rbind(blank_years, fish)
```

### Plotting

At this point, its time to plot the data. I took some time to play around with `annotate()` this week - it provides some interesting use for labeling data in a more customized format.

```{r Plotting}
fishplot = fish %>%
  ggplot(aes(x = ifelse(category=="sustainable",yes=-share, no=share), 
             y = as.factor(year), 
             fill = category)
            ) +
  geom_col(show.legend = FALSE) + 
  scale_x_continuous(labels = abs, breaks = c(-91,0,34), limits = c(-92,50)) +
  scale_fill_manual(values = c("white", "#FC766AFF", "#5B84B1FF")) +
  theme_alex() + 
  xlab("Share of Fish Population") +
  ylab("") +
  geom_vline(xintercept=0, color = "white", linetype = "dashed")+
  theme(
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_blank(),
    plot.title=element_text(hjust = 0),
    plot.subtitle=element_text(hjust = 0),
    plot.caption=element_text(hjust = 1)
  ) +
  annotate(geom="text",x = 40, y = "1974", color = "white", label = "1974") +
  annotate(geom="text",x = 40, y = "1980", color = "white", label = "1980") +
  annotate(geom="text",x = 40, y = "1990", color = "white", label = "1990") +
  annotate(geom="text",x = 40, y = "2000", color = "white", label = "2000") +
  annotate(geom="text",x = 40, y = "2010", color = "white", label = "2010") +
  annotate(geom="text",x = 40, y = "2017", color = "white", label = "2017") +
  annotate(geom="text",x =-40, y = "2002", color = "white", label = "Sustainable", size = 6) +
  annotate(geom="text",x = 20, y = "2002", color = "white", label = "Overexploited", size = 6) +
  labs(title = "Sustainabile Fishing over Time",
       subtitle = "Overexplotation of fish populations has increased over time",
       caption = "#TidyTuesday | Data: OurWorldInData.org \n @unstandarderror")

fishplot
```


